cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.Rmd
Updated S-curve data generation functions

The `gen_quadratic(n, p, range)` function generates a dataset of $n$ points forming a quadratic curve in the first two dimensions of a $p$-dimensional space. This $2\text{-}D$ parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The `gen_cubic(n, p, range)` function generates a dataset of $n$ points forming a cubic curve in the first two dimensions of a $p$-dimensional space. This function creates a more complex curvilinear structure than a simple parabola.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function `gen_scurve(n, p)`, which generates `n` observations in `p` dimensions.

The $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-\frac{3\pi}{2}, \frac{3\pi}{2}\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure:

$$
  X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta) \cdot (\cos(\theta) - 1).
$$

This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction.

For $p > 3$, additional noise dimensions are appended introducing structured, wavy perturbations.

@@

To simulate a variant of the S-curve structure with a removed region ("hole"), we define the function `gen_scurvehole(n, p)`. This function produces `n` observations in a $p$-dimensional space where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

To simulate missing regions on the manifold, a fixed anchor point $(0, 1, 0, \ldots)$ is defined in $p$-dimensional space. All observations within a Euclidean distance of $\sqrt{0.3} \approx 0.5477$ from the anchor are removed:

$$
\text{remove if} \quad \sum_{j=1}^p (X_j - a_j)^2 \leq 0.3
$$.

This exclusion creates a hole in the manifold centered near the middle vertical region of the S-curve.

>>>>

The `gen_quadratic(n, p, range)` function generates a dataset of $n$ points forming a quadratic curve in the first two dimensions. This $2\text{-}D$ parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The `gen_cubic(n, p, range)` function generates a dataset of $n$ points forming a cubic curve in the first two dimensions. This function creates a more complex curvilinear structure than a simple parabola.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function `gen_scurve(n, p)`, which generates $n$ observations in $p\text{-}D$.

The $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-\frac{3\pi}{2}, \frac{3\pi}{2}\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$, $X_3 = \text{sign}(\theta) \cdot (\cos(\theta) - 1)$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction. For $p > 3$, additional noise dimensions are appended introducing structured, wavy noise.

@@

To simulate a variant of the S-curve structure with a removed region ("hole"), we define the function `gen_scurvehole(n, p)`. This function produces $n$ observations in $p\text{-}D$ where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

To simulate missing regions on the manifold, a fixed anchor point $(0, 1, 0, \ldots)$ is defined in $p\text{-}D$. All observations within a Euclidean distance of $\sqrt{0.3} \approx 0.5477$ from the anchor are removed if $\sum_{j=1}^p (X_j - a_j)^2 \leq 0.3$. This exclusion creates a hole in the manifold centered near the middle vertical region of the S-curve.

### 2025-06-24 paper-cardinalR.tex
Updated S-curve data generation functions

The \texttt{gen\_quadratic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a quadratic curve in the first two dimensions of a \(p\)-dimensional space. This \(2\text{-}D\) parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The \texttt{gen\_cubic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a cubic curve in the first two dimensions of a \(p\)-dimensional space. This function creates a more complex curvilinear structure than a simple parabola.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function \texttt{gen\_scurve(n,\ p)}, which generates \texttt{n} observations in \texttt{p} dimensions.

The \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-\frac{3\pi}{2}, \frac{3\pi}{2}\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure:

\[
  X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta) \cdot (\cos(\theta) - 1).
\]

This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction.

For \(p > 3\), additional noise dimensions are appended introducing structured, wavy perturbations.

@@

To simulate a variant of the S-curve structure with a removed region (``hole''), we define the function \texttt{gen\_scurvehole(n,\ p)}. This function produces \texttt{n} observations in a \(p\)-dimensional space where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

To simulate missing regions on the manifold, a fixed anchor point \((0, 1, 0, \ldots)\) is defined in \(p\)-dimensional space. All observations within a Euclidean distance of \(\sqrt{0.3} \approx 0.5477\) from the anchor are removed:

\[
\text{remove if} \quad \sum_{j=1}^p (X_j - a_j)^2 \leq 0.3
\].

This exclusion creates a hole in the manifold centered near the middle vertical region of the S-curve.

>>>>

The \texttt{gen\_quadratic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a quadratic curve in the first two dimensions. This \(2\text{-}D\) parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The \texttt{gen\_cubic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a cubic curve in the first two dimensions. This function creates a more complex curvilinear structure than a simple parabola.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function \texttt{gen\_scurve(n,\ p)}, which generates \(n\) observations in \(p\text{-}D\).

The \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-\frac{3\pi}{2}, \frac{3\pi}{2}\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\), \(X_3 = \text{sign}(\theta) \cdot (\cos(\theta) - 1)\). This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction. For \(p > 3\), additional noise dimensions are appended introducing structured, wavy noise.

@@

To simulate a variant of the S-curve structure with a removed region (``hole''), we define the function \texttt{gen\_scurvehole(n,\ p)}. This function produces \(n\) observations in \(p\text{-}D\) where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

To simulate missing regions on the manifold, a fixed anchor point \((0, 1, 0, \ldots)\) is defined in \(p\text{-}D\). All observations within a Euclidean distance of \(\sqrt{0.3} \approx 0.5477\) from the anchor are removed if \(\sum_{j=1}^p (X_j - a_j)^2 \leq 0.3\). This exclusion creates a hole in the manifold centered near the middle vertical region of the S-curve.
